# Submodel structure \label{sec:submod}

The `a4a` stock assessment framework allows the user to set up a large number of different models. The mechanics which provide this flexibility are designed around the concept of submodels. The user has to define the model structure using `R` formulas, including `mgcv` [@R-mgcv] gam formulas, for each unknown variable that must be estimated. By using `R` formulas the stock assessment framework gives lot's of flexibility to explore models and combinations of submodels.

There are 5 submodels in operation:

- a model for F-at-age ($F_{ay}$)
- a (list) of model(s) for abundance indices catchability-at-age ($Q_{ays}$),
- a model for recruitment ($R_y$)
- a list of models for the observation variance of catch-at-age and abundance indices ($\{\sigma^2_{ay}, \tau^2_{ays}\}$)
- a model for the initial age structure ($N_{a,y=1}$)

When setting the structure of each submodel the user is in fact building the predictive model and its parameters. The optimization process, done through `ADMB`, estimates the parameters and their variance-covariance matrix, allowing further analysis to be carried out, like simulation, prediction, diagnostics, etc. All the statistical machinery will be at the user's reach.

```{R, message=FALSE, warning=FALSE, echo=FALSE}
# libraries
library(FLa4a)
# datasets
data(ple4)
data(ple4.indices)
```

## Submodel building blocks and fundamental `R` formulas

The elements available to build submodels formulas are 'age' and 'year', which can be used to build models with different structures. 

In R's modelling language, a constant model is coded as $\sim 1$, while a slope over time would simply be $\sim year$, a smoother over time $\sim s(year, k=10)$, a model with a coefficient for each year would be $\sim factor(year)$. Transformations of the variables are as usual, e.g. $\sim sqrt(year)$, etc; while combinations of all the above can be done although non-convergence will limit the possibilities.

Using the $F$ submodel as example the following specifies the models described in the previous paragraph, see Figure \@ref(fig:fundforms).

```{r, fundforms, fig.cap="Example of fundamental R formulas used to model the year effect"}
# models
m1 <- ~1
m2 <- ~ year
m3 <- ~ s(year, k=10)
m4 <- ~ factor(year)
m5 <- ~ sqrt(year)

# fits
fit1 <- sca(ple4, ple4.indices, fmodel=m1, fit="MP")
fit2 <- sca(ple4, ple4.indices, fmodel=m2, fit="MP")
fit3 <- sca(ple4, ple4.indices, fmodel=m3, fit="MP")
fit4 <- sca(ple4, ple4.indices, fmodel=m4, fit="MP")
fit5 <- sca(ple4, ple4.indices, fmodel=m5, fit="MP")

# plot
lst <- FLStocks(constant=ple4+fit1,
  linear=ple4+fit2,
  smooth=ple4+fit3,
  factor=ple4+fit4,
  sqrt=ple4+fit5)
lst <- lapply(lst, fbar)
lgnd <- list(points=FALSE, lines=TRUE, space='right')
xyplot(data~year, groups=qname, lst, auto.key=lgnd, type='l', ylab='fishing mortality')
```

The models above and their combinations can be used to model both 'age' and 'year'. The corresponding fits for age are show in the code below and Figure \@ref(fig:fundformsage).

```{r, fundformsage, fig.cap="Example of fundamental R formulas  used to model the age effect"}
# models
m1 <- ~1
m2 <- ~ age
m3 <- ~ s(age, k=3)
m4 <- ~ factor(age)
m5 <- ~ sqrt(age)

# fits
fit1 <- sca(ple4, ple4.indices, fmodel=m1, fit="MP")
fit2 <- sca(ple4, ple4.indices, fmodel=m2, fit="MP")
fit3 <- sca(ple4, ple4.indices, fmodel=m3, fit="MP")
fit4 <- sca(ple4, ple4.indices, fmodel=m4, fit="MP")
fit5 <- sca(ple4, ple4.indices, fmodel=m5, fit="MP")

# plot
lst <- FLStocks(constant=ple4+fit1,
  linear=ple4+fit2,
  smooth=ple4+fit3,
  factor=ple4+fit4,
  sqrt=ple4+fit5)
lst <- lapply(lst, function(x) harvest(x)[,'2000'])
xyplot(data~age, groups=qname, lst, auto.key=lgnd, type='l', ylab='fishing mortality in 2000')
```

## The major effects available for modelling

Although the building blocks for formulas are 'age' and 'year', in fact there are three effects that can be modelled for each submodel: 'age', 'year' and 'cohort'. As examples note the following models for fishing mortality.

```{r}
# the age effect
ageeffect <- ~ factor(age)

# the year effect
yeareffect <- ~ factor(year)

# the cohort
cohorteffect <- ~ factor(year-age)

# the fits
fit1 <- sca(ple4, ple4.indices, fmodel=yeareffect)
fit2 <- sca(ple4, ple4.indices, fmodel=ageeffect)
fit3 <- sca(ple4, ple4.indices, fmodel=cohorteffect)
```

and the graphical representation of the three models in Figures \@ref(fig:majeffy) to \@ref(fig:majeffc).

```{r, majeffy, fig.cap="Major effects: the year effect (~ factor(year))", echo=FALSE}
wireframe(harvest(fit1), main='year effect')
```

```{r, majeffa, fig.cap="Major effects: the age effect (~ factor(age))", echo=FALSE}
wireframe(harvest(fit2), main='age effect')
```

```{r, majeffc, fig.cap="Major effects: the cohort effect (~ factor(year-age))", echo=FALSE}
wireframe(harvest(fit3), main='cohort effect')
```

## The submodel class and methods

Although the specification of each submodel is done through a R formula, internally the `a4a` sca fit creates an object (of class 'submodel') which stores more information and allows a wide number of methods to be applied. The most important cases are prediction and simulation methods.

The submodel class is a S4 class with the following slots:

```{r}
showClass("submodel")
```

Objects of class 'submodel' are created in the fitting process using the formulas set by the user (or defaults if missing) and information from the fit. 

For example begining with a simple `a4a` fit:

```{r}
 # fit a model with indices "IBTS_Q1" and "IBTS_Q3"
 fit0 <- sca(ple4, ple4.indices[c("IBTS_Q1", "IBTS_Q3")],
             fmodel = ~ s(age, k = 5),
             qmodel = list( ~ s(age, k = 4), ~ s(age, k = 4)),
             srmodel = ~ 1,
             n1model = ~ s(age, k = 5),
             vmodel = list( ~ 1, ~ 1, ~ 1),
             verbose = FALSE)
```

```{r, echo=FALSE}
  fit0
```

