---
title: "Untitled"
author: "Danai"
date: "2024-10-27"
output: html_document
---

# Getting help for S4 classes and methods

```{r, echo=FALSE, message=FALSE}
library(FLCore)
library(FLa4a)
library(ggplotFL)
```
`a4a` is build around S4 classes. S4 classes and methods in R offer an object-oriented programming framework but in order to access the documentation requires specific terminology. In this section we will demonstrate how to get information on the main building blocks of `a4a`.

For example, `FLStock` is one of our main components in order to run our stock assessment model. We can check the structure of an `FLStock` object as follows:

```{r}
showClass("FLStock")
```

The object oriented structure of `a4a` gives the opportunity to change the behavior of a function according to the object that is applied to. For example we can check the available methods of the function `plot`

```{r}
showMethods("plot")
```

by calling `showMethods` R prints all the possible uses of the **plot** function. We want to see what it does when it is called on an `FLStock` object with no other object. We observe that `plot` takes two arguments, `x` and `y`. So, in the signature of the `getMethod` function we are going to use, we need to define both `x` and `y`.

```{r}
getMethod('plot', signature = list("FLStock","missing"))
```


