# Submodel structure \label{sec:submod}

The `a4a` stock assessment framework allows the user to set up a large number of different models. The mechanics which provide this flexibility are designed around the concept of submodels. Each unknown variable that must be estimated is treated as a linear model, for which the user has to define the model structure using `R` formulas, including [`mgcv`](http://cran.r-project.org/web/packages/mgcv/index.html) gam formulas. All submodels use the same specification process, the `R` formula interface, wich gives lot's of flexibility to explore models and combination of submodels.

There are 5 submodels in operation:

- a model for F-at-age ($F_{ay}$)
- a (list) of model(s) for abundance indices catchability-at-age ($Q_{ays}$),
- a model for recruitment ($R_y$)
- a list of models for the observation variance of catch-at-age and abundance indices ($\{\sigma^2_{ay}, \tau^2_{ays}\}$)
- a model for the initial age structure $N_{a,y=1}$,

When setting the structure of each submodel the user is in fact building the predictive model and its parameters. The optimization process, done through `ADMB`, estimates the parameters and their variance-covariance matrix, allowing further analysis to be carried out, like simulation, prediction, diagnostics, etc. All the statistical machinery will be at the user's reach.

## Submodel building blocks and fundamental `R` formulas

The elements available to build submodels formulas are 'age' and 'year', which can be used to build models with different structures. 

In R's linear modelling language, a constant model is coded as $ 1$, while a slope over time would simply be $\sim year$, a smoother over time $\sim s(year, k=10)$, a model with a coefficient for each year would be $\sim factor(year)$. Transformations of the variables are as usual, e.g. $\sim sqrt(year)$, etc; while combinations of all the above can be done although non-convergence will limit the possibilities. 

Using the $F$ submodel as example the following specifies the models described in the previous paragraph:

```{r, fund_forms, fig.cap="Example of fundamental R formulas"}
# models
m1 <- ~1
m2 <- ~ year
m3 <- ~ s(year, k=10)
m4 <- ~ factor(year)
m5 <- ~ sqrt(year)

# fits
fit1 <- sca(ple4, ple4.indices, fmodel=m1, fit="MP")
fit2 <- sca(ple4, ple4.indices, fmodel=m2, fit="MP")
fit3 <- sca(ple4, ple4.indices, fmodel=m3, fit="MP")
fit4 <- sca(ple4, ple4.indices, fmodel=m4, fit="MP")
fit5 <- sca(ple4, ple4.indices, fmodel=m5, fit="MP")

# plot
lst <- FLStocks(constant=ple4+fit1, linear=ple4+fit2, smooth=ple4+fit3, factor=ple4+fit4, sqrt=ple4+fit5)
lst <- lapply(lst, fbar)
lgnd <- list(points=FALSE, lines=TRUE, space='right')
xyplot(data~year, groups=qname, lst, auto.key=lgnd, type='l', ylab='fishing mortality')
```

The models above and their combinations can be used to model both 'age' and 'year'. The corresponding fits for age are:

```{r, fund_forms_age, fig.cap="Example of fundamental R formulas"}
# models
m1 <- ~1
m2 <- ~ age
m3 <- ~ s(age, k=3)
m4 <- ~ factor(age)
m5 <- ~ sqrt(age)

# fits
fit1 <- sca(ple4, ple4.indices, fmodel=m1, fit="MP")
fit2 <- sca(ple4, ple4.indices, fmodel=m2, fit="MP")
fit3 <- sca(ple4, ple4.indices, fmodel=m3, fit="MP")
fit4 <- sca(ple4, ple4.indices, fmodel=m4, fit="MP")
fit5 <- sca(ple4, ple4.indices, fmodel=m5, fit="MP")

# plot
lst <- FLStocks(constant=ple4+fit1, linear=ple4+fit2, smooth=ple4+fit3, factor=ple4+fit4, sqrt=ple4+fit5)
lst <- lapply(lst, function(x) harvest(x)[,'2000'])
xyplot(data~age, groups=qname, lst, auto.key=lgnd, type='l', ylab='fishing mortality in 2000')
```

## The major effects available for modelling

Although the building blocks for formulas are 'age' and 'year', in fact there are three effects that can be modelled for each submodel: 'age', 'year' and 'cohort'. As examples note the following models for fishing mortality.

```{r}
# the age effect
ageeffect <- ~ factor(age)

# the year effect
yeareffect <- ~ factor(year)

# the cohort
cohorteffect <- ~ factor(year-age)

# the fits
fit1 <- sca(ple4, ple4.indices, fmodel=yeareffect)
fit2 <- sca(ple4, ple4.indices, fmodel=ageeffect)
fit3 <- sca(ple4, ple4.indices, fmodel=cohorteffect)
```

and the graphical representation of the three models in Figures \@ref(fig:majeffy) to \@ref(fig:majeffc).

```{r, majeffy, fig.cap="Major effects: the year effect (~ factor(year))"}
wireframe(harvest(fit1), main='year effect')
```

```{r, majeffa, fig.cap="Major effects: the age effect (~ factor(age))"}
wireframe(harvest(fit2), main='age effect')
```

```{r, majeffc, fig.cap="Major effects: the cohort effect (~ factor(year-age))"}
wireframe(harvest(fit3), main='cohort effect')
```

## The submodel class and methods

Although the specification of each submodel is done through a R formula, internally the `a4a` sca fit creates an object (of class 'submodel') which stores more information and allows a wide number of methods to be applied. The most important cases are prediction and simulation methods.

The submodel class is a S4 class with the following slots:

```{r}
showClass("submodel")
```

Objects of class 'submodel' are created in the fitting process using the formulas set by the user (or defaults if missing) and information from the fit. 

For example begining with a simple `a4a` fit:

```{r}
 # fit a model with indices "IBTS_Q1" and "IBTS_Q3"
 fit0 <- sca(ple4, ple4.indices[c("IBTS_Q1", "IBTS_Q3")],
             fmodel = ~ s(age, k = 5),
             qmodel = list( ~ s(age, k = 4), ~ s(age, k = 4)),
             srmodel = ~ 1,
             n1model = ~ s(age, k = 5),
             vmodel = list( ~ 1, ~ 1, ~ 1),
             verbose = FALSE)
```

```{r, echo=FALSE}
  fit0
```

[SECTION NEEDS REVISION, WE NOW HAVE STKMODEL]
Within the sca fit object there's a collection of submodels for each component of the `a4a` stock assessment model (fishing mortality, survey catchability, stock recruitment relationship, initial population and the observation variance for catch numbers and survey indices). For readability the following example will use the fishing mortality submodel although the methods and models described can be applied to each of the 5 submodels.

The submodel can be extracted using:

```{r,}
 fmod <- fmodel(fit0)
```

inside this object is a formula, the coeficients estimated by the model and their covariance matrix.

```{r,sim_pred_fmodel_coef, results = 'hide', eval=FALSE}
 coef(fmod)
 vcov(fmod)
```

it is also possible to get the data behind the fit and the design matrix required to get the fitted values


[COLIN] do we still have the getX method?


```{r,results = 'hide', eval=FALSE}
 as.data.frame(fmod)
 getX(fmod)
```

this makes it possble to check the fit manually like this (note that the data is centered in the model, which is addressed by adding the centering after the prediction is made, see code below):

```{r, eval=FALSE} 
 dat <- as.data.frame(fmod)
 X <- getX(fmod)
 b <- coef(fmod)
 dat$fit <- exp(c(X %*% b))# + dat$data.centering)

 plot(fit ~ age, type = "l", data = dat[dat$year == 2016,],
      ylab = "Estimated fishing mortality at age",
      ylim = c(0, max(dat$fit)), las = 1)
```

or simulate from the fit by resampling based on the variance matrix of the coefficients, which is done by simulating many coefficients (here called `bsim`).

```{r, eval=FALSE}
 # get the variance matrix of the coefficients and simulate
 Sigma <- vcov(fmod)[,,1]
 bsim <- mvrnorm(999, b, Sigma)

 fit_sim <- exp(X %*% bsim)
 dat$ci_lower <- apply(fit_sim, 1, quantile, p = 0.025)
 dat$ci_upper <- apply(fit_sim, 1, quantile, p = 0.975)

 plot(fit ~ age, type = "l", data = dat[dat$year == 2016,],
      ylab = "Estimated fishing mortality at age",
      ylim = c(0, max(dat$ci_upper)), las = 1)
 lines(ci_lower ~ age, data = dat[dat$year == 2016,], lty = 2)
 lines(ci_upper ~ age, data = dat[dat$year == 2016,], lty = 2)
```

The above code is to show how it is possible to use an `FLa4a` submodel to predict and simulate, and not intended to be used by the user.  The recomended way to predict from a submodel and make simulations is to use the `genFLQuant` function (and for finer control, there is also a `simulate` function). The above example can be done using `FLa4a` functions as follows:

```{r, eval=FALSE}
 fmod_fit <- genFLQuant(fmod)
 dat <- as.data.frame(fmod_fit)

 plot(data ~ age, type = "l", data = dat[dat$year == 2016,],
      ylab = "Estimated fishing mortality at age",
      ylim = c(0, max(dat$data)), las = 1)
```
or using the `ggplotFL` package:

```{r, eval=FALSE}
 fmod_fit <- genFLQuant(fmod)

 ggplot(data = fmod_fit[,"2016"], aes(x = age, y = data)) + 
   geom_point() + geom_line() + 
   ylab("Estimated fishing mortality at age") +
   ylim(0, max(fmod_fit))
```

and the simulations and confidence intervals are easily computed:

```{r,  eval=FALSE}
 # simulate 999 
 fmod_fit_sim <- genFLQuant(fmod, nsim = 999)

 # reduce to quantiles
 fmod_fit_sim <- quantile(fmod_fit_sim[,"2016"], c(0.025, 0.50, 0.975))
```

plotting can be done by converting to a data.frame, reshaping and using standard `ggplot2` functionality

```{r,  eval=FALSE}
 dat <- 
   reshape(
     as.data.frame(fmod_fit_sim, drop=TRUE), 
     timevar = "iter", idvar = "age", direction = "wide"  
   )

 ggplot(data=dat, aes(x = age, y = `data.50%`)) +
   geom_ribbon(aes(ymin = `data.2.5%`, ymax = `data.97.5%`), 
               fill="red", alpha = .15) +
   geom_point() + geom_line() + 
   ylab("Estimated fishing mortality at age") +
   ylim(0, max(fmod_fit_sim))
```


