# Predict and simulate \label{sec:predsim}


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(mgcv)
library(ggplot2);theme_set(theme_bw())
library(FLCore)
library(FLa4a)
library(ggplotFL)
library(ggpubr)
drafting <- FALSE
```

To predict and simulate `R` uses the methods `predict()` and `simulate()`, which are implemented in `FLa4a`, adapted to work with several of the model classes.  The simulate and predict methods only work if the mode object has be run with the option `fit = "assessment"`, and not `fit = "MP"`, the later is used when space and time are at a premium, for example in a Management Strategy Evaluation. In this chapter we will only cover prediction and simulation of full stock objects.

This chapter is based on the following model:
```{r, predsim_fit0}
fit0 <- sca(ple4, ple4.index)
submodels(fit0)
```

Simulation and prediction in FLa4a is based on two functions: `simulate` and `predict`.

## predict()

Predict simply computes the quantities of interest using the estimated coefficients and the design matrix of the model, defined via the formulas in the submodels. In the case of predicting from a fitted model, a list is returned, with one element for each submodel, and each element of the list is a `FLQuants` object with the following `FLQuant`s

```{r, predsim_fit_pred}
fit.pred <- predict(fit0)
lapply(fit.pred, names)
```

This allows easy access to the parameterised parts of the model, for example the initial population structure, `ny1`, can be accessed via `fit.pred$stkmodel$ny1`. If the whole stock is of interest, for example, to inspect model predictions of SSB, then the following code can be used

```{r, predsim_stk_pred}
stk.pred <- ple4 + fit0
```

The `+` operator is used to add the predictions from the model to the initial stock object. Then quantities can be extracted as normal, `ssb(stk.pred)`.


## simulate()

Simulations are always done by generating random draws from a multivariate normal distribution with mean given by the coeffients of the model, and variance matrix given by the estimated covariance matrix of the coefficients (in practice this is a submatrix of the inverse of the hessian matrix, see details below).

simulate performed on a fitted model returns a fitted model with the same structure as the original model, but with the model parameters replaced by `nsim` simulated parameters. The following code simulates the model 200 times, and a histogram of the result show in Figure \@ref(fig:predsim_hist).

```{r, predsim_fit_sim}
sim.fit <- simulate(fit0, nsim = 200)
```

```{r, predsim_hist, fig.cap="Histogram of 200 draws from the approximate distribution of the estimate of survey observation error."}
hist(
  exp(coef(sim.fit)$vmodel$IND),
  main = "200 draws of a model parameter",
  nclass = 20,
  xlab = "observation std error for survey index"
)
```


The fitted model can also be used to generate some predictions, such as `harvest`, `stock.n` or `catch.n`. Figure \@ref(fig:sim) shows a comparison between the estimated values and the medians of the simulation, while Figure \@ref(fig:sim2) presents the stock summary of the simulated and fitted data.

```{r}
fits <- simulate(fit0, 100)
flqs <- FLQuants(sim=iterMedians(stock.n(fits)), det=stock.n(fit0))
```

```{r, sim, fig.cap="Median simulations VS fit"}
xyplot(data~year|age, groups=qname, data=flqs, type="l",
       scales=list(y=list(relation="free", draw=FALSE)),
       auto.key=list(points=FALSE, lines=TRUE, columns=2),
       par.settings=list(superpose.line=list(col=c("gray35", "black")),
       strip.background=list(col="gray90")), ylab="")
```

```{r, sim2, fig.cap="Stock summary of the simulated and fitted data"}
stks <- ple4 + fits
plot(stks)
```



## simulate() and predict()

The `simulate()` and `predict()` functions can be combined to simulate and then predict from each of the simulations. The following code simulates the model 100 times and then predicts the fitted quantitites from the model, resulting in simulations from the core model elements: initial age structure, recruitment, fishing mortality, survey catchability and observation variance. A plot of the initial age structure is shown as an example in Figure \@ref(fig:sim_ny1).

```{r}
sim.pred <- predict(sim.fit)
```

```{r, sim_ny1, fig.cap='Simulations from the model prediction of initial age structure', fig.pos = 'H', fig.height = 4,echo=FALSE, message=FALSE, warning=FALSE}

fit_sim_ny1 <- sim.pred$stkmodel$ny1[-1, ]

# reduce to quantiles
fit_sim_ny1 <- quantile(fit_sim_ny1, prob = c(0.025, 0.50, 0.975))

# reshape
dat <- reshape(
  as.data.frame(fit_sim_ny1, drop = TRUE),
  timevar = "iter", idvar = c("age"), direction = "wide"
)

# plot
ggplot(data = dat, aes(x = age, y = `data.50%`)) +
  geom_ribbon(aes(ymin = `data.2.5%`, ymax = `data.97.5%`),
    fill = "red", alpha = .15
  ) +
  geom_point() +
  geom_line() +
  ylab("Estimated initial age structure (numbers)") +
  scale_x_continuous(breaks = pretty(dat$age))
```

If interest is in the complete stock, then the `+` operator can be used to add the simulated quantities to the initial stock object, as shown below. The stock information can be extracted, such as SSB, a plot of which is shown in Figure \@ref(fig:sim_ssb).

```{r}
stk.sim <- ple4 + sim.fit
```

```{r, sim_ssb, fig.cap='Simulations from the model prediction of initial age structure', fig.pos = 'H', fig.height = 4,echo=FALSE, message=FALSE, warning=FALSE}

fit_sim_ny1 <- ssb(stk.sim)

# reduce to quantiles
fit_sim_ny1 <- quantile(fit_sim_ny1, prob = c(0.025, 0.50, 0.975))

# reshape
dat <- reshape(
  as.data.frame(fit_sim_ny1, drop = TRUE),
  timevar = "iter", idvar = c("year"), direction = "wide"
)

# plot
ggplot(data = dat, aes(x = year, y = `data.50%`)) +
  geom_ribbon(aes(ymin = `data.2.5%`, ymax = `data.97.5%`),
    fill = "red", alpha = .15
  ) +
  geom_point() +
  geom_line() +
  ylab("Estimated spawning stock biomass") +
  scale_x_continuous(breaks = pretty(dat$year))
```

## Details

Simulate uses the variance-covariance matrix computed from the Hessian returned by `ADMB` and the fitted parameters, to parametrize a multivariate normal distribution. The simulations are carried out using the method `mvrnorm()` provided by the R package [MASS](http://cran.r-project.org/web/packages/MASS/).  Such an approach is often called a parametric bootstrap, and is a common method for generating uncertainty in the parameters of a model. It is imporatant to not that the method approximated the joint distribution of the model parameters as a multivariate normal, which is true if the model is linear and well specified.  Since stock assessment models tend to be non-linear, a more robust approach is to use MCMC methods instread.

Another point to note, is that in order to save space, only the  subset of the covariance matrix is stored for each submodel. This in effect sets any correlations between submodel parameters to zero in simulations.
