# Predict and simulate \label{sec:predsim}


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(mgcv)
library(ggplot2);theme_set(theme_bw())
library(FLCore)
library(FLa4a)
library(ggplotFL)
library(ggpubr)
drafting <- FALSE
```

To predict and simulate `R` uses the methods `predict()` and `simulate()`, which are implemented in `FLa4a`, adapted to work with several of the model classes.  The simulate and predict methods only work if the mode object has be run with the option `fit = "assessment"`, and not `fit = "MP"`, and the later is used when space and time are at a premium, for example in a Management Strategy Evaluation.

This chapter is based on the following model:
```{r, pred0_fit0}
fit0 <- sca(ple4, ple4.index)
submodels(fit0)
```

## Basic functions

Simulation and prediction in FLa4a is based on two functions: `simulate` and `predict`.

### simulate()

Unlike the stats function `stats::simulate`, `FLa4a::simulate` will return the same object as it was passed. For example, if you simulate from a FLa4a fit, you will get an FLa4a fit object back, in which the coefficients of the object are simulations from the model. Likewise, if you call simulate on a submodel object you will get back a submodel object inwhich the coefficients are simulations from the model. Simulations are always done by generating random draws from a multivariate normal distribution with mean given by the coeffients of the model, and variance matrix given by the estimated covariance matrix of the coefficients (in practice this is a submatrix of the inverse of the hessian matrix).

Simulate works on several classes from full fits right down to the individual model compents, so if `my\_fit` is a fitted `a4a` model, then `simulate(fit0, nsim = 100)` will return a new fitted `a4a` model where the model coefficients now have 100 iters and are drawn from the full variance matrix of the fitted model. Similarly, `simulate(stkmodel(my_fit), nsim = 100)` will return a stock submodel with the same formulas as fitted model as in the `a4a` fit but where the coefficients are simulated from the variance matrix of the relavent parameters.  See also, `simulate(qmodel(my_fit), nsim = 100)` to simulate from the survey catchability submodels.

## stock submodel

In an `sca` fit individual submodel objects are often combined into a collection of `submodels`, for example the models for survey catchability are a collection of submodels

```{r, sim_pred_qsubmodels}
stk_submodel <- stkmodel(fit0)
```

which contains a submodel for each survey catchability. Now (almost) the same code can be run as before to plot the estimates with confidence intervals, the difference with submodels is all the results will be `FLQuants` so, `lapply` must be used to do computations on the predictions for each submodel seperately. 

```{r, sim_pred_qmodel_predict_a4a}
qmod_fit_sim <- genFLQuant(qmod, nsim = 999)

# reduce to quantiles
qmod_fit_sim <- lapply(qmod_fit_sim, "[", j = "2016")
qmod_fit_sim <- lapply(qmod_fit_sim, quantile, prob = c(0.025, 0.50, 0.975))

# reshape
dat <- reshape(
    as.data.frame(qmod_fit_sim, drop=TRUE), 
    timevar = "iter", idvar = c("age", "qname"), direction = "wide"  
  )

# plot
ggplot(data=dat, aes(x = age, y = `data.50%`)) +
  geom_ribbon(aes(ymin = `data.2.5%`, ymax = `data.97.5%`), 
              fill = "red", alpha = .15) +
  geom_point() + geom_line() +
  ylab("Estimated catchability at age") +
  facet_wrap(~qname, scales = "free_x") +
  scale_x_continuous(breaks = pretty(dat$age))
```

the coefficients and variance covariance are all available via

```{r, results = "hide", eval = FALSE}
coef(qmod)
vcov(qmod)
```

## Predict

Predict simply computes the quantities of interest using the estimated coefficients and the design matrix of the model.

```{r}
fit.pred <- predict(fit0)
lapply(fit.pred, names)
```

## Simulate

Simulate uses the variance-covariance matrix computed from the Hessian returned by `ADMB` and the fitted parameters, to parametrize a multivariate normal distribution. The simulations are carried out using the method `mvrnorm()` provided by the R package [MASS](http://cran.r-project.org/web/packages/MASS/). Figure \@ref(fig:sim) shows a comparison between the estimated values and the medians of the simulation, while Figure \@ref(fig:sim2) presents the stock summary of the simulated and fitted data.

```{r}
fits <- simulate(fit0, 100)
flqs <- FLQuants(sim=iterMedians(stock.n(fits)), det=stock.n(fit0))
```

```{r, sim, fig.cap="Median simulations VS fit"}
xyplot(data~year|age, groups=qname, data=flqs, type="l",
       scales=list(y=list(relation="free", draw=FALSE)),
       auto.key=list(points=FALSE, lines=TRUE, columns=2),
       par.settings=list(superpose.line=list(col=c("gray35", "black")),
       strip.background=list(col="gray90")), ylab="")
```

```{r, sim2, fig.cap="Stock summary of the simulated and fitted data"}
stks <- ple4 + fits
plot(stks)
```

## Simulate and Predict

The `simulate()` and `predict()` functions can be combined to simulate and then predict from each of the simulations. The following code simulates the model 100 times and then predicts the fitted quantitites from the model, resulting in simulations from the core model elements: initial age structure, recruitment, fishing mortality, survey catchability and observation variance. A plot of the initial age structure is shown as an example.

```{r}
fit_sim <- predict(simulate(fit0, 100))
```

```{r, fig.cap='Generated data with two linear fits, left a simple mean and right a straight line', fig.pos = 'H', fig.height = 4,echo=FALSE, message=FALSE, warning=FALSE}
fit_sim_ny1 <- fit_sim$stkmodel$ny1[-1,]

# reduce to quantiles
fit_sim_ny1 <- quantile(fit_sim_ny1, prob = c(0.025, 0.50, 0.975))

# reshape
dat <- reshape(
  as.data.frame(fit_sim_ny1, drop = TRUE),
  timevar = "iter", idvar = c("age"), direction = "wide"
)

# plot
ggplot(data = dat, aes(x = age, y = `data.50%`)) +
  geom_ribbon(aes(ymin = `data.2.5%`, ymax = `data.97.5%`),
    fill = "red", alpha = .15
  ) +
  geom_point() +
  geom_line() +
  ylab("Estimated initial age structure (numbers)") +
  scale_x_continuous(breaks = pretty(dat$age))
```
