# Projections and harvest control rules

Massad et al. (2018) provide a useful distinction between forecasting and projection in scientific prediction, stating that prediction in general science can be categorized into two main components: forecasting and projection. According to their definition, "a forecast is an attempt to predict what will happen, whereas a projection describes what would happen, given certain hypotheses" (Massad et al., 2018). This distinction is also reflected in the Organisation for Economic Co-operation and Development (OECD, 2009), which notes that “forecasting” and “prediction” are often used interchangeably in assessing the magnitude that a given variable will assume in the future.

In the context of economic and environmental modeling, the OECD distinguishes between medium-term projections, which typically extend five to six years into the future, and other predictive exercises that rely on different methodologies. The term "projection" is generally used in two interrelated senses: (1) as a future value of a time series computed based on specific assumptions about environmental changes, and (2) in probability theory, as the conditional expectation of a variable. Since regression models provide conditional expectations of dependent variables based on predictor variables, this probabilistic use of projection aligns with forecasting and predictive modeling (OECD, 2009).

Applying this conceptual framework to fisheries science, the process of advising on future fishing opportunities can be viewed as a combination of forecasting and projection. When using a harvest control rule (HCR) approved by decision-makers to predict catch levels that can be extracted from a stock, the analysis falls into the realm of forecasting, as it predicts what will happen under known system conditions. Conversely, when assessing potential future outcomes based on predefined scenarios—such as climate change effects or policy interventions—the approach aligns more closely with projections.

In fisheries science, the distinction between forecasting and prediction is not always explicitly made, with the discipline generally emphasizing broader predictive approaches. It is common practice to run scenarios when advising fisheries managers, testing different assumptions about future ecological and management conditions. Strictly speaking, projections are not part of the stock assessment process itself. Stock assessment concludes when analysts compare estimates of biomass and fishing mortality with reference points, allowing for determinations of whether a stock is overfished or subject to overfishing (Hilborn & Walters, 1992).

Projections typically follow stock assessments, incorporating estimates or assumptions about population dynamics—growth, reproduction, and natural mortality—to predict future catches, biomass, and abundance under specific conditions and with quantified uncertainty. By integrating both forecasting and projection methodologies, fisheries science can offer robust, scenario-based decision-making tools that aid sustainable resource management (Punt et al., 2016).

### References:
- Hilborn, R., & Walters, C. J. (1992). *Quantitative fisheries stock assessment: Choice, dynamics and uncertainty*. Chapman & Hall.
- Massad, E., Coutinho, F. A. B., Lopez, L. F., & da Silva, D. R. (2018). *Mathematical models for infectious diseases: A theoretical companion*. Springer.
- OECD. (2009). *The OECD economic outlook: Sources and methods*. Organisation for Economic Co-operation and Development.
- Punt, A. E., Smith, A. D. M., & Cui, G. (2016). Review of progress in the introduction of management strategy evaluation (MSE) approaches in fisheries management. *Fish and Fisheries*, 17(1), 235–265.

For this section we'll be using the package `FLasher` (REF) from the FLR family of packages.

```{r, echo=FALSE}
library(FLa4a)
library(FLasher)
library(FLBRP)
library(ggplotFL)
data(ple4)
data(ple4.indices)
```

## Simple workflow

The basic workflow to project with `FLasher` is to extend the `FLStock` object to store the predictions using the method `fwdWindow`, set the targets for the projection with method `fwdControl` and project the fishery with the method `fwd` with a `FLStock` and a `FLSR`.

We'll start by fitting a model including, a stock recruitment model which will be used to forecast recruitment. We'll also set the number of iterations we'll be working with and the time period we want to project.

```{r, warning=FALSE, message=FALSE}
# fit model
fit00 <- sca(ple4, ple4.indices, srmodel=~geomean()) 
stk00 <- ple4 + fit00
# create stock recruitment model object
sr00 <- as(fit00, "FLSR")
# set projection
# number of iterations
nsim <- 250
# most recent year in the data
maxy <- range(ple4)["maxyear"]
# number of years to project
projy <- 5
# last year for projections
endpy <- maxy + projy
# initial year for projections
inipy <- maxy + 1
# extend stock object to store projection's results
stk00 <- fwdWindow(stk00, end = endpy)
# set the control for the projections, in this case a fixed f of 0.3
trg00 <- fwdControl(year = inipy:endpy, quant = "f", value = 0.3)
# project
stk01 <- fwd(stk00, control=trg00, sr=sr00)
```

```{r, echo=FALSE, fig.cap="Projection of stock for 5 years with fixed fishing mortality and recruitment"}
plot(stk01)
```

A natural addition to this forecast is to add uncertainty. We'll do that by generating uncertainty in population numbers, catch numbers and fishing mortality, using `simulate`, and add stock recruitment uncertainty using the residuals of the fit. 

```{r, warning=FALSE, message=FALSE}
stk00 <- ple4 + simulate(fit00, nsim)
stk00 <- fwdWindow(stk00, end = endpy)
res00 <- residuals(sr00)
rec00 <- window(rec(stk00), 2018, 2022)
rec00 <- rlnorm(rec00, mean(res00), sqrt(var(res00)))
stk02 <- fwd(stk00, control=trg00, sr=sr00, residuals=rec00)
```

```{r, echo=FALSE, fig.cap="Stochastic projection of stock for 5 years with fixed fishing mortality and recruitment"}
plot(stk02)
```

An alternative to the above workflow is to fit the stock recruitment model after the stock assessment model, using the output of the assessment as input to the stock recruitment fit. In which case stock recruitment estimation uncertainty can be added by fitting the stock recruitment model over stock assessment uncertainty, so that there will be stock-recruitment fit to each iteration generated from the stock assessment model.

```{r, warning=FALSE, message=FALSE}
fit00 <- sca(ple4, ple4.indices) 
stk00 <- ple4 + simulate(fit00, nsim)
sr00 <- as.FLSR(stk00, model="geomean")
sr00 <- fmle(sr00, control = list(trace = 0))
stk00 <- fwdWindow(stk00, end = endpy)
res00 <- residuals(sr00)
rec00 <- window(rec(stk00), inipy, endpy)
rec00 <- rlnorm(rec00, c(yearMeans(res00)), sqrt(c(yearVars(res00))))
stk03 <- fwd(stk00, control=trg00, sr=sr00, residuals=rec00)
```
[IAGO, check rlnorm, it doesn't work if the meanlog and sdlog are FLQuants]


```{r, echo=FALSE, fig.cap="Stochastic projection of stock for 5 years with fixed fishing mortality and recruitment"}
plot(stk03)
```

These two methods don't give very different result when the stock recruitment model is not having a large impact in the other parameters. However the second method is much slower due to all the fits needed to have the empirical distribution of the stock recruitment model parameters.

```{r, echo=FALSE, fig.cap="Stochastic projection of stock for 5 years with fixed fishing mortality and recruitment. 01: projection without uncertainty, stock recruitment model fit within the stock assessment model; 02: projection with uncertainty, stock recruitment model fit within the stock assessment model; 03: projection with uncertainty, stock recruitment model fit after the stock assessment model"}
plot(window(FLStocks('01' = stk01, '02' = stk02, '03' = stk03), start = 2000))
```

## Initial condition assumptions [CHECK fwdWindow]

When projecting the stock forward one needs to make a number of assumptions about initial conditions, the starting point from where projections will be made. The method `fwdWindow` has a set of options that allows the analyst to decide about those assumptions:

- `wt`: Number of years to average over to get the future mean weights at age, default is 3
- `mat`: Number of years to average over to get the future proportion mature at age, default is 3
- `m`: Number of years to average over to get the future natural mortality at age, default is 3
- `spwn`: Number of years to average over to get the future fraction of mortality before spawning, default is 3
- `discards.ratio`: Number of years to average over to get the future mean proportion of discards at age. Default is 3
- `catch.sel`: Number of years to average over to get the future selection patern (fishing mortality at age which will be scaled based on canges in $\bar{F}$). Default is 3

One can also define if those assumptions will be based on the mean value over the time period set, or randomly sampled from historical values, through setting the argument `fun` to `mean` or `sample`, respectively.

For the next examples we'll use the approach of fitting the stock recruitment within the assessment together with other parameters. We'll set to 20 the number of years to compute mean weights at age, to 10 the number of years to average across and estimate the selection pattern in terms of fishing mortality at age. Finally, we'll use a 10 year period to compute the average discard ratio.

```{r, warning=FALSE, message=FALSE}
fit00 <- sca(ple4, ple4.indices, srmodel=~geomean()) 
sr00 <- as(fit00, "FLSR")
stk00 <- ple4 + fit00
stk00 <- fwdWindow(stk00, end = endpy, years = list(wt = 20, catch.sel = 10, discards.ratio = 10), fun = list(wt = "sample"))
trg00 <- fwdControl(year = inipy:endpy, quant = "f", value = 0.3)
stk04 <- fwd(stk00, control=trg00, sr=sr00)
```

```{r, echo=FALSE, fig.cap="Stochastic projections of stock for 5 years with fixed fishing mortality and recruitment. Two scenarios with different assumptions about initial conditions"}
plot(FLStocks(default=stk01, alternative=stk04))
```

## Scenarios

There's a wide range of scenarios that can be of interest to project in order to give advice to policy makers, or to better understand the fitted stock assessment model. For example, projecting the stock in the absence of fishing for a few generations, gives good insights about the dynamics of the population being modelled.

```{r, warning=FALSE, message=FALSE}
fit00 <- sca(ple4, ple4.indices, srmodel=~geomean())
sr00 <- as(fit00, "FLSR")
stk00 <- ple4 + simulate(fit00, nsim)
# set projection
projy <- 25
endpy <- maxy + projy
inipy <- maxy + 1
stk00 <- fwdWindow(stk00, end = endpy)
trg00 <- fwdControl(year = inipy:endpy, quant = "f", value = 0)
# recruitment uncertainty
res00 <- residuals(sr00)
rec00 <- window(rec(stk00), inipy, endpy)
rec00 <- rlnorm(rec00, mean(res00), sqrt(var(res00)))
# project
stk05 <- fwd(stk00, control=trg00, sr=sr00, residuals=rec00)
```

```{r, echo=FALSE, fig.cap="Stochastic projection of stock for 25 years in the absence of fishing"}
plot(stk05)
```

These scenarios are defined by the target quantities one's trying to achieve. In `FLasher` there are the following target quantities:

- srp
- ssb_end
- biomass_end
- ssb_spawn
- biomass_spawn
- ssb_flash
- biomass_flash
- inmb_end
- indb
- catch
- landings
- discards
- f
- fbar
- revenue
- effort

[IAGO NEED YOUR HELP HERE]

When projecting the stock under the conditions defined by the scenario one can mix several quantities. For example it may be interesting to project an initial situation of growing the stock followed by a higher exploitation to evaluate how catches would behave.

```{r, warning=FALSE, message=FALSE}
trg00 <- fwdControl(year = inipy:endpy, quant = c(rep("ssb_end", 15), rep("f", 10)), value = c(rep(2000000, 15), rep(0.3, 10)))
stk06 <- fwd(stk00, control=trg00, sr=sr00, residuals=rec00)
```

```{r, echo=FALSE, fig.cap="Stochastic projection of stock for 25 years with fixed SSB for 15 years followed by fixed fishing mortality for 10 years and constant recruitment"}
plot(stk06)
```

## Relative scenarios

Another scenario that is very useful when advising decision makers is to have objectives which are relative to previous preformances. For example one could increase spwanwing stock biomass by 10% each year. This is done buy using the argument `relYear` and setting `value` in relative terms, $1.1$.

```{r, warning=FALSE, message=FALSE}
fit00 <- sca(ple4, ple4.indices, srmodel=~geomean())
sr00 <- as(fit00, "FLSR")
stk00 <- ple4 + simulate(fit00, nsim)
# set projection
projy <- 5
endpy <- maxy + projy
inipy <- maxy + 1
stk00 <- fwdWindow(stk00, end = endpy)
trg00 <- fwdControl(year = inipy:endpy, quant = "ssb_end", value = 1.1, relYear = inipy:endpy-1)
# recruitment uncertainty
res00 <- residuals(sr00)
rec00 <- window(rec(stk00), inipy, endpy)
rec00 <- rlnorm(rec00, mean(res00), sqrt(var(res00)))
# project
stk07 <- fwd(stk00, control=trg00, sr=sr00, residuals=rec00)
```

Similar scenarios can be set for all quantities and any years to use as reference. The next example sets a scenario where $SSB$ levels are set in relation to the most recent estimate out of the assessmeent.

```{r, warning=FALSE, message=FALSE}
trg00 <- fwdControl(year = inipy:endpy, quant = "ssb_end", value = 1.1, relYear = maxy)
stk08 <- fwd(stk00, control=trg00, sr=sr00, residuals=rec00)
```

```{r, echo=FALSE, fig.cap="Stochastic projection of stock for 25 years with fixed SSB for 15 years followed by fixed fishing mortality for 10 years and constant recruitment"}
plot(window(FLStocks("10% SSB growth relative to previous year" = stk07, "10% higher SSB relative to most recent estimate" = stk08), start = 2000))
```

## Limits

An important element when projecting the stock forward is to keep the performance of the fishery within some boundaries. A common one requested by the industry is to keep catches within some stability. `fwd` can include those constraints using the `min` and `max` arguments. The next example sets the minimum future catches to half of mean historical catches.

```{r, warning=FALSE, message=FALSE}
minc <- 0.2*mean(catch(stk00), na.rm=TRUE)
trg00 <- fwdControl(year = inipy:endpy, quant = rep(c("ssb_end", "catch"), projy), value = rep(c(1500000, NA), projy), min=rep(c(NA, minc), projy))
stk09 <- fwd(stk00, control=trg00, sr=sr00, residuals=rec00)
```

```{r, echo=FALSE, fig.cap="Stochastic projection of stock for 25 years with SSB target of 1500000t and catch limit of 50% historical catches"}
plot(window(stk09, start = 2000))
```

## Harvest Control Rule

Harvest Control Rules (HCR) can be complex and of many shapes (REF). We'll keep our examples simple to demonstrate the mechanism of coding HCR.

HCR are decision algorythms that can be used to codify the decision making process, allowing for longer term stability of management decisions in fisheries.

The HCR we're going to explore is based on a target and a limit. The target is applied to the management objective and represents the intent of management. The limit is applied to the process we want to use as trigger for protective actions. For example, the objective of the management system is to extract the highest catches possible for a very long time (aka equilibrium). However, due to natural variability and scientific uncertainty, it can happen that the stock's biomass decreases below what's expected, in which case decision makers want to make sure the stock stays healthy and productive. This situation can be translated into a target of fishing mortality at the level that extracts the Maximum Sustainable Yield, and a biomass limit of half the biomass that would produce the referred catches, everything being in a stable equilibrium. If SSB falls below the limit then fishing mortality is set at 80% of the target.

Such HCR could be written as

$if SSB_y > B_{MSY} then F_{y+1} = F_{MSY} or else F_{y+1} = 0.9xF_{MSY}$

and coded like

```{r, warning=FALSE, message=FALSE}
# # fit model
# fit00 <- sca(ple4, ple4.indices, srmodel=~geomean())
# stk00 <- ple4 + fit00
# # create stock recruitment model object
# sr00 <- as(fit00, "FLSR")
# # estimate reference points
# #brp00 <-
# #ftrg <- fmsy
# #blim <- ssb
# # set projection
# # most recent year in the data
# maxy <- range(ple4)["maxyear"]
# # number of years to project
# projy <- 2
# # last year for projections
# endpy <- maxy + projy
# # initial year for projections
# inipy <- maxy + 1
# # extend stock object to store projection's results
# stk00 <- fwdWindow(stk00, end = endpy)
# # set the controls for the projections
# trg00 <- fwdControl(year = inipy:endpy, quant = "f", value = trg)
# trg01 <- fwdControl(year = inipy:endpy, quant = "f", value = 0.8trg)
# # project
# if(ssb(stk00)>blim) stk10 <- fwd(stk00, control=trg00, sr=sr00) elseif stk10 <- fwd(stk00, control=trg01, sr=sr00)
```
















