# Reference Points

One of the primary objectives of stock assessment is the estimation of reference points. These serve as benchmarks for evaluating the outputs of assessment models and determining the status of a fish stock. Reference points are critical for effective fisheries management, supporting decision makers setting future levels of exploitation.

The most common classification of stock status is bidimensional, comparing exploitation levels and biomass sizes against target reference points. This framework allows for the assessment of whether a stock is overfished or experiencing overfishing:

1. Exploitation Levels: Typically represented by fishing mortality ($F$), overfishing occurs when $F$ exceeds the target reference point.
   
2. Biomass Size: Commonly measured by spawning stock biomass ($SSB$), stocks are deemed overfished if the $SSB$ falls below the reference point.

These assessments often utilize tools like the Kobe plot, which visually represents stock status in relation to these metrics (REF).

In addition to target reference points, limit reference points (LRPs) are commonly included in stock assessments. These represent thresholds that should not be crossed, as they signal a high risk of stock collapse or significant uncertainty in population dynamics. Effective management aims to maintain fishing pressure and biomass levels away from these limits to ensure long-term sustainability and reduce the risk of adverse outcomes.


In 1982, the United Nations Convention on the Law of the Sea (UNCLOS, @unclos1982), in Article 61, required coastal states to determine and maintain, or restore, populations of harvested species at levels capable of producing the maximum sustainable yield (MSY). MSY is defined as the largest yield (catch) that can be taken from a specific fish stock over an indefinite period without causing stock depletion. Since then MSY and its proxies, such as $B_{MSY}$ (biomass at MSY) and $F_{MSY}$ (fishing mortality at MSY), remain widely used. These reference points are related with the stock's productivity, which results of a complex interaction between recruitment, individual growth and mortality processes.

Recruitment refers to the addition of new individuals to a fish population, specifically those that become vulnerable to fishing. It includes the process of spawning, which depends on the reproductive potential of adults, and the survival of larvae until they enter the fishery - a process largely influenced by environmental conditions. Individual growth determines how long it takes for a fish to gain weight, increase in length, and eventually reach maturity and reproduce. Mortality is typically divided into natural mortality and fishing mortality. Natural mortality includes all causes of death unrelated to fishing, such as predation by other species, and is influenced by both environmental factors and biological interactions. Fishing mortality, on the other hand, is primarily driven by human activity and depends on decisions related to fishing practices, including the effort exerted by the fleet, the selectivity of the gear used, and the availability of fish. For example, the productivity of a stock can differ significantly depending on whether a fleet fishes in an area with many juvenile fish using small-mesh nets, or in an area where juvenile fish are scarce and large-mesh nets are used. These processes are interconnected and influenced by a complex mix of environmental conditions and biological characteristics.

For this section we'll be using the package `FLBRP` [@flbrp] from the FLR family of packages and its documentation for parts of the text.

```{r}
library(FLBRP)
library(FLa4a)
data(ple4)
data(ple4.indices)
fit0 <- sca(ple4, ple4.indices)
stk0 <- ple4 + fit0
```

To proceed with the computation of reference points we must start by creating an `FLBRP` object and afterwards run the fitting process with `brp()`. The `FLBRP` class has information on selection pattern, mass at age, and biological parameters (see `?FLBRP` for a full description of this class).

Slots named `*.obs` will contain the related time series present in the original `FLStock` object, while other slots will contain averages across the year dimension over the last $n$ years, where $n$ is controlled by the arguments `biol.nyears`, `fbar.nyears` and `sel.nyears` (Table \@ref(tab:brpargs))

| Argument | Default value | Description |
|---:|:---|:---------|
|fbar | seq(0, 4, length.out = 101)| Vector of Fs to find minimum or maximum value of metric, e.g. MSY |
| nyears | 3 | Number of years to average across to estimate input quantities for reference points estimation |
| biol.nyears | nyears | Number of years to average across to estimate biological quantities, e.g. maturity|
| fbar.nyears | nyears | Number of years to average across to scale F-at-age to 1 and compute selection pattern|
| sel.nyears | nyears | Number of years to average across to compute landings and discards in number of individual per age |
| na.rm | TRUE | Remove NA if existing|
| mean | "arithmetic" | Statistic to average quantities above, alternative is "geometric" for the geometric mean|

Table: (\#tab:brpargs) Arguments available to create `FLBRP` object.

By default `FLBRP` creates a harvest slot with 100 computations of fishing mortality at age scaled from $\bar{F}=0$ up to $F_{crash}$ or $\bar{F}=4$, if the former isn't possible to compute, which is later used to find the reference points.

## Yield per recruit reference points

In the case where no stock recruitment relationship exists, or was fitted, `brp()` will return yield per recruit reference points. By default it computes biomasses in the absence of fishing, also know as virgin biomass, $F_{MAX}$, $F_{0.1}$ and 40% Spawning per recruit reference points.

```{r}
brp0 <- FLBRP(stk0)
brp0 <- brp(brp0)
summary(brp0)
```

The selection pattern and other quantities can be depicted by calling `plot()` on the specific `FLBRP` object's slot.

```{r, selection pattern, echo=FALSE}
xyplot(data~age,data=catch.sel(brp0),type=c('l', 'p'))
```

```{r, relevant quantities for reference points, echo=FALSE}
xyplot(data~age|qname, data=FLQuants(sel=catch.sel(brp0),
 	dsel=discards.sel(brp0), swt=stock.wt(brp0),
	cwt =catch.wt(brp0), mat= mat(brp0), m = m(brp0)),
	type="l",scale="free")
```

To extract a table with all reference points one uses the method `refpts()`. Note in this case $F_{msy}$ is the same as $F_{max}$, since the assumed stock recruitment is mean recruitment. 

```{r}
refpts(brp0)
refpts(brp0)[c('msy', ('fmax')), ]
```

The depiction of the reference points with the method `plot()` shows recruitment as constant over all levels of biomass and set to $1$.

```{r, reference points, echo=FALSE}
plot(brp0)
```

## Stock recruitment relationship based reference points

An important way to improve reference points is to include stock recruitment dynamics. Yield per recruit, as in the previous section, ignores these dynamics and assumes recruitment will be the same no matter $SSB$'s size. To inform `brp()` to take stock recruitment dynamics into account, the stock recruitment model must be fitted and the resulting `FLSR` object passed to the `FLBRP` call when creating the `FLBRP` object.

There's two ways of fitting stock recruitment models: (i) after fitting the stock assessment model by using its outputs, $SSB$ and recruitment, as data to fit the model; (ii) inside the stock assessment model together with all other quantities. There's pros and cons on both approaches, we're not going to dwell on those now though.

### Stock recruitment after fitting the stock assessment model  

In the following example we'll use a Beverton and Holt stock recruitment relationship. There are several other relationships that can be used, see `?bevholt` for more details.

```{r, message=FALSE, warning=FALSE, message=FALSE}
sr0 <- as.FLSR(stk0, model=bevholt)
sr0 <- fmle(sr0, control = list(trace = 0))
plot(sr0)
```

We now need to provide the `FLSR` object, `sr0`, to the `FLBRP()` and refit the reference points.

```{r}
brp0 <- FLBRP(stk0, sr=sr0)
model(brp0)
params(brp0)
brp0 <- brp(brp0)
```

The new reference points can now be extracted using `refpts()` with the `FLBRP` object as the main argument, and depict the relationships with `plot()`. Note this time by setting the flag `obs` to `TRUE` the plot will include the estimates of $SSB$ and $R$.

```{r}
refpts(brp0)
```

Note $MSY$ based reference points are no longer the same as $F_{MAX}$, and recruitment is no longer constant over all $SSB$ levels.

```{r, echo=FALSE}
plot(brp0, obs=TRUE)
```

### Stock recruitment during stock assessment model fit

An alternative option, using `sca()`, its to fit the stock recruitment model together with the stock assessment model fit, and create the `FLSR` object from the fit object (Figure \@ref(fig:a4asr)).

```{r}
# fit with Beverton and Holt model
fit1 <- sca(ple4, ple4.indices, srmodel = ~ bevholt(CV = 0.5))
# create FLSR object
a4aflsr <- as(fit1, "FLSR")
```

```{r, a4asr, fig.cap="Stock recruitment model estimated by `sca()`"}
plot(a4aflsr, obs = TRUE)
```

Create the `FLBRP` object with the new `FLSR` object with the stock recruitment model fitted with `sca()` and fit a new set of reference points (Figure \@ref(fig:a4arp)). Note the reference points are slightly different. The stock recruitment parameters estimated with `sca()` take into account all the other parameters and as such are not exactly the same, ultimately resulting in a distinct set of reference points.

```{r}
# create FLBRP object
a4abrp <- FLBRP(stk0, a4aflsr)
a4abrp <- brp(a4abrp)
```

```{r, a4arp, fig.cap="Reference points estimated with `sca()` stock recruitment fit"}
plot(a4aflsr, obs = TRUE)
```

## Economics reference points

We can add economic data to the `FLBRP` object to calculate economic based reference points, like maximum economic yield (MEY). We need to provide information about price, variable costs and fixed costs. The first in value at age per weight of fish, the others in value per unit of fishing mortality.

```{r}
# price
price(brp0) <- c(rep(1,3),rep(1.5,2),rep(2,5))
price(brp0)@units <- "1000 euro per ton"

# variable costs per F 
vcost(brp0) <- 100000
vcost(brp0)@units <- "1000 euro per F"

# fixed costs per F 
fcost(brp0) <- 50000
fcost(brp0)@units <- "1000 euro per F"

# reference points
brp0 <- brp(brp0)
refpts(brp0)
```

The reference points table is now complete with values for `revenue`, `costs` and `profit`, as well as estimtes for $MEY$ based reference points. The point where `profits` are maximized, instead of the point where `catch` is maximized as in the case of `MSY` (Figure \@ref(fig:ecorp)).

```{r ecorp, fig.cap="Reference points including economic reference points"}
plot(brp0)
```

## Computing user specific reference points

The user may want to calculate specific "reference points" given F levels. The example below shows how it can be done, having in mind that by specifying F levels the user may be computing arbitrary references.

```{r}
custom_refs <- FLPar(Ftrgt1 = 0.33, Ftrgt2 = 0.44)
brp1 <- brp0 + custom_refs
refpts(brp1)
```

<!--Or create an empty FLPar with specified reference points and recalculate everything:

```{r}
#brp2 <- FLPar(NA,dimnames=list(refpt=c("virgin","f0.1","fmax","spr.30","spr.35","spr.45"), quantity=c("harvest","yield","rec","ssb","biomass","revenue","cost","profit"), iter=1))
brp2 <- brp1

brp2@refpts <- FLPar(NA, dimnames=list(refpt = c("virgin", "f0.1", "fmax", "spr.30", "spr.35","spr.45"), quantity=c("harvest", "yield", "rec", "ssb", "biomass", "revenue", "cost", "profit"), iter=1))

brp2 <- brp(brp2)
refpts(brp2)
```
Another simple way, although it onl;y works for $SPR$ based reference points, is to include other `spr.##` points in the `refpts` table.
-->

One specific case of user tailored reference points is to compute $F_{MSY}$ ranges according to @HILBORN2010193 and @Rindorf_etal_2016 ideas. For this case there's already the method `msyRanges()`, which takes as argument a fitted `FLBRP` object and delivers a `FLPar` object, similar to `refpts` with the lower and upper boundary of F according to the specified range multiplier, e.g. if 0.05 the ranges will reflect $(1-0.05) \times F_{MSY}$ and $(1+0.05) \times F_{MSY}$.

```{r}
rp.rngs <- msyRange(brp0, range=0.05)
rp.rngs
```

## Assessing the status of the stock

One of the most important outcomes of the stock assessment process is determining the status of the stock relative to the objectives for $SSB$ and $F$.

On one hand, these objectives reflect the level of risk policy makers are willing to accept, expressed as biomass levels intended to remain at sea, typically linked to the replenishing capacity of the stock and risk of collapse. For example, larger biomasses generally results in lower probability of collapse, increased recruitment and stock replenishing.

On the other hand, objectives consider food production, represented by the fishing mortality levels ($F$). At similar $SSB$ levels, higher fishing mortality will result in larger catches but will also cause greater reductions in $SSB$ after the fishing season.

Policy makers, with the support of scientific advisors, have to find a balance between these two objectives, food production and stock collapse, to guarantee the long term sustainability of the stocks. As such computing the status of the stock in relation to $SSB$ and $F$ becomes one of the most important pieces of information scientists can provide.

For this exercise we'll use the kobe plot (REF), which depicts estimates of $SSB$ over $SSB_{target}$ and $F$ over $F_{target}$ in a plot with $SSB$ in the y axis and $F$ in the x axis. The inspection of the plot allows a quick overview of the exploitation history of the stock and the current status in relation to the objectives set.

```{r}
# code to compute kobe plot
```



