# Annex - stock assessment workflow

[WHY? HOW!]

## The "mean" model

To start the analysis we'll fit a "mean" model, where all submodels will be set to an overall average, by using the $\sim 1$ formula. This will be our reference model to see how adding age and year effects will show up in the diagnostic tools, in particular in the residuals.

```{r}
library(FLa4a)
data(hke1567)
data(hke1567.idx)
fit01 <- sca(hke1567, hke1567.idx, fmod=~1, qmod=list(~1), srmod=~1, vmod=list(~1, ~1),  n1mod=~1)
res01 <- residuals(fit01, hke1567, hke1567.idx)
```

The common residuals plot clearly shows a trend across ages (Figure \@ref(fig:meanresbyage)) for both datasets.

```{r, meanresbyyear, fig.cap="Mean fit residuals by year)"}
plot(res01)
```

Which is even clearer when plotting the residuals by age across years.

```{r, meanresbyage, fig.cap="Mean fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res01, auxline="l", by="age")
```

## The age effects

These models will introduce age effects in the fishing mortality submodel and catchability submodel. First in the fishinf mortality submodel.

```{r}
fit02 <- sca(hke1567, hke1567.idx, fmod=~factor(age), qmod=list(~1), srmod=~1, vmod=list(~1, ~1),  n1mod=~1)
res02 <- residuals(fit02, hke1567, hke1567.idx)
```

The residuals plot now shows catch at age residuals less stagered, reflecting the modelling of the age effect.

```{r, fageresbyyear, fig.cap="f age effect fit residuals by year)"}
plot(res02)
```

The residuals plot by age shows the same outcome.

```{r, fageresbyage, fig.cap="f age effect fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res02, auxline="l", by="age")
```

Follwed by the same addition to the catchability model.

```{r}
fit03 <- sca(hke1567, hke1567.idx, fmod=~1, qmod=list(~factor(age)), srmod=~1, vmod=list(~1, ~1),  n1mod=~1)
res03 <- residuals(fit03, hke1567, hke1567.idx)
```

```{r, qageresbyyear, fig.cap="q age effect fit residuals by year)"}
plot(res03)
```

The residuals plot by age shows the same outcome.

```{r, qageresbyage, fig.cap="q age effect fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res03, auxline="l", by="age")
```

Finally both effects are brought together.

```{r}
fit04 <- sca(hke1567, hke1567.idx, fmod=~factor(age), qmod=list(~factor(age)), srmod=~1, vmod=list(~1, ~1),  n1mod=~1)
res04 <- residuals(fit04, hke1567, hke1567.idx)
```

```{r, fqageresbyyear, fig.cap="q age effect fit residuals by year)"}
plot(res04)
```

The residuals plot by age shows the same outcome.

```{r, fqageresbyage, fig.cap="q age effect fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res04, auxline="l", by="age")
```

## The fishing mortality year model

This model will introduce an year effect in the fishing mortality submodel on top of the F age effect added before.

```{r}
fit05 <- sca(hke1567, hke1567.idx, fmod=~factor(age) + factor(year), qmod=list(~1), srmod=~1, vmod=list(~1, ~1),  n1mod=~1)
res05 <- residuals(fit05, hke1567, hke1567.idx)
```

The residuals plot now shows catch at age residuals stagered as before. The year trends are less pronounced although, because the data doesn't have a very strong year effect, it's less clear than when modelling the age effect.

```{r, fyearresbyyear, fig.cap="f year effect fit residuals by year)"}
plot(res05)
```

The residuals plot by age shows the same outcome.

```{r, fyearresbyage, fig.cap="f year effect fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res05, auxline="l", by="age")
```

We can see now that the residuals show a lot less patterns than before. There's still some issues, the survey catchability seems to have an year trend. However the model is not fully specified yet, stock recruitment is modelled as constant over time, the initial population abundance is also modelled as a constant as well as the variance models.

### The initial year population abundance model, aka N1

This model will introduce an age effect in the population abundance in the first year of the time series. This model sets the n-at-age in the first year of the time series, which is needed due to the lack of previous data to reconstruct those cohorts.

```{r}
fit06 <- sca(hke1567, hke1567.idx, fmod=~factor(age) + factor(year), qmod=list(~factor(age)), srmod=~1, vmod=list(~1, ~1),  n1mod=~factor(age))
res06 <- residuals(fit06, hke1567, hke1567.idx)
```

The residuals plot now shows catch at age residuals stagered as before. The year trends are less pronounced although, because the data doesn't have a very strong year effect, it's less clear than when modelling the age effect.

```{r, n1arresbyyear, fig.cap="f year effect fit residuals by year)"}
plot(res06)
```

The residuals by age (Figure \@ref(fig:n1resbyage)) the residuals' improvement in the first year of the catch at age time series (bottom left plots).

```{r, n1resbyage, fig.cap="f year effect fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res06, auxline="l", by="age")
```

## The stock recruitment submodel

In this example we'll simply add a model to allow recruitment to vary over time and we'll see how to track potential improvements in the residuals.

```{r}
fit07 <- sca(hke1567, hke1567.idx, fmod=~factor(age) + factor(year), qmod=list(~factor(age)), srmod=~factor(year), vmod=list(~1, ~1),  n1mod=~factor(age))
res07 <- residuals(fit07, hke1567, hke1567.idx)
```

The residuals plot by year are very useful to see the effect of adding a varying stock recruitment model. The year trends present in previous models are not absent. Recruitment variability when left unmodelled was being picked up by trends in the survey catchability and catch at age. And due to the cohort dynamics underlying the catch at age model, where propagating into other ages' estimates.

```{r, srresbyyear, fig.cap="f year effect fit residuals by year)"}
plot(res07)
```

```{r, srresbyage, fig.cap="f year effect fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res07, auxline="l", by="age")
```

## The variance submodel

Finally, we're testing the variance submodel, specifically the catch at age variance model. We won't dig into the catchability variance model though. It's common to accept that a scientific survey following a well designed sampling protocol will have equal variance across ages since no preferential areas are sampled.

```{r}
fit10 <- sca(hke1567, hke1567.idx, fmod=~factor(age) + s(year, k=10), qmod=list(~factor(age)), srmod=~s(year, k=10), vmod=list(~factor(age), ~1),  n1mod=~factor(age))
res10 <- residuals(fit10, hke1567, hke1567.idx)
```

To see what's happening with the variance model one can use predict to plot the different models fitted.

```{r, vagepredbyage, fig.cap="Variance models for catch at age"}
flqs <- FLQuants(mod10=predict(fit10)$vmodel$catch[,"2022"], mod07=predict(fit07)$vmodel$catch[,"2022"])
xyplot(data~age, data=flqs, group=qname, type="l", auto.key=T)
```

To see the effect these models have on the estimated quantities one can look at the variance of the estimates:

```{r, vage, fig.cap="Estimates of population abundance with different variance models"}
flqs <- FLQuants(mod10=catch.n(hke1567+simulate(fit10, nsim=500))[,"2022"], mod07=catch.n(hke1567+simulate(fit07, nsim=500))[,"2022"])
bwplot(data~qname|factor(age),  data=as.data.frame(flqs), scales="free", auto.key=T)
```

and the usual residuals

```{r, vageresbyyear, fig.cap="f year effect fit residuals by year)"}
plot(res10, auxline="r")
```

```{r, vageresbyage, fig.cap="f year effect fit residuals by age)",  out.extra = "angle=90", out.height="18cm", out.width="25cm", fig.height=8, fig.width=12}
plot(res10, auxline="l", by="age")
```
