# Annex - stock assessment workflow

The following sections describes a potential workflow for fitting a `a4a` stock assessment model. The idea is to explore the age and year effects in isolation and adjust the model's smoothness to model those effects.

The procedure is heavily supported by residuals' analysis. In a well specified model residuals should show a random pattern, without any trend or very high values (outliers).

## The "mean" model

To start the analysis we'll fit a "mean" model, where all submodels will be set to an overall average, by using the `~1` formula. This will be our reference model to see how adding age and year effects will show up in the diagnostic tools, in particular in the residuals.

```{r, message=FALSE, warning=FALSE}
library(FLa4a)
data(hke1567)
data(hke1567.idx)
fit01 <- sca(hke1567, hke1567.idx,
    fmod=~1,
    qmod=list(~1),
    srmod=~1,
    vmod=list(~1, ~1),
    n1mod=~1)
res01 <- residuals(fit01, hke1567, hke1567.idx)
```

The common residuals plot clearly shows a time trend for each age (Figure \@ref(fig:meanresbyage)) for both datasets. Furthermore, inspecting how catch ate age residuals are positioned across ages, by comparing the level of residuals for each age, one can see the pattern of lower than 0 residuals in age 0, reversing the signal for ages 1 and 2, close to 0 in ages 3 and 4, and again below 0 in age 5.


```{r, meanresbyyear, fig.cap="Mean fit residuals by year)"}
plot(res01)
```

This pattern becomes more apparent when plotting the residuals by age across years.

```{r, meanresbyage, fig.cap="Mean fit residuals by age"}
plot(res01, auxline="l", by="age")
```

## The age effects

The following models will introduce age effects in the fishing mortality submodel and catchability submodel. In the fishing mortality submodel we'll introduce a `factor` which means that there will be as many parameters as ages minus 1 and each parameters will be independent of each other.

```{r}
fit02 <- sca(hke1567, hke1567.idx,
    fmod=~factor(age),
    qmod=list(~1),
    srmod=~1,
    vmod=list(~1, ~1),
    n1mod=~1)
res02 <- residuals(fit02, hke1567, hke1567.idx)
```

The residuals plot now shows catch at age residuals less staggered, reflecting the modelling of the age effect.

```{r, fageresbyyear, fig.cap="Fishing mortality model with age effect residuals by year"}
plot(res02)
```

The residuals plot by age shows the same outcome.

```{r, fageresbyage, fig.cap="Fishing mortality model with age effect residuals by age"}
plot(res02, auxline="l", by="age")
```

We'll now proceed adding an age effect to the catchability model while removing the catch at age effect.

```{r}
fit03 <- sca(hke1567, hke1567.idx,
    fmod=~1,
    qmod=list(~factor(age)),
    srmod=~1,
    vmod=list(~1, ~1),
    n1mod=~1)
res03 <- residuals(fit03, hke1567, hke1567.idx)
```

```{r, qageresbyyear, fig.cap="Index catchability model with age effect residuals by year"}
plot(res03)
```

The residuals plot by age shows the same outcome.

```{r, qageresbyage, fig.cap="Index catchability model with age effect residuals by age"}
plot(res03, auxline="l", by="age")
```

Finally both effects are brought together.

```{r}
fit04 <- sca(hke1567, hke1567.idx,
    fmod=~factor(age),
    qmod=list(~factor(age)),
    srmod=~1,
    vmod=list(~1, ~1),
    n1mod=~1)
res04 <- residuals(fit04, hke1567, hke1567.idx)
```

```{r, fqageresbyyear, fig.cap="Fishing mortality and index catchability models with age effect residuals by year"}
plot(res04)
```

The residuals plot by age shows the same outcome.

```{r, fqageresbyage, fig.cap="Fishing mortality and index catchability models with age effect residuals by age"}
plot(res04, auxline="l", by="age")
```

## Year effect on fishing mortality

This model will introduce an year effect in the fishing mortality submodel on top of the F age effect added before. Inspecting the last set of residuals (Figure \@ref(fig:fqageresbyage)) one can easily see the pattern across years with more positive residuals in the beginning of the time series and more negative in the most recent years. As for age we're using a `factor` for years. The new model's residuals won't show such a pronounced effect anymore (Figures \@ref(fig:fyearresbyyear) and \@ref(fig:fyearresbyage)).

```{r}
fit05 <- sca(hke1567, hke1567.idx,
    fmod=~factor(age) + factor(year),
    qmod=list(~1),
    srmod=~1,
    vmod=list(~1, ~1),
    n1mod=~1)
res05 <- residuals(fit05, hke1567, hke1567.idx)
```

```{r, fyearresbyyear, fig.cap="Fishing mortality model with year effect residuals by year"}
plot(res05)
```

The residuals plot by age shows the same outcome.

```{r, fyearresbyage, fig.cap="Fishing mortality model with year effect residuals by age"}
plot(res05, auxline="l", by="age")
```

We can see now that the residuals show a lot less patterns than before. There's still some issues, the survey catchability seems to have an year trend. However the model is not fully specified yet, stock recruitment is modelled as constant over time, the initial population abundance is also modelled as a constant as well as the variance models.

## Year effect on catchability

It's uncommon to include year trends on the abundance index catchability model. Such decision needs to be considered carefully as the trend in the index, in the case of a well design scientific survey, should result from a change in abundance. Modelling that trend would attribute such change to the survey design and remove it from the abundance. If the survey index is based on a commercial CPUE it becomes more likely that changes in selectivity or fishing behaviour could show up in the index as changes in abundance. Although the common process of standardizing CPUEs should deal with technical issues.

In the case of adding year effects to the catchability submodel the same formulas can be used, to include period breaks, trends, etc.

## The initial year population abundance model, aka N1

This model sets the n-at-age in the first year of the time series, which is needed due to the lack of previous data to reconstruct those cohorts. It will affect the population numbers in the lower triangle of the initial population matrix and catches.

The following model will introduce an age effect in the population abundance in the first year of the time series.

```{r}
fit06 <- sca(hke1567, hke1567.idx,
    fmod=~factor(age) + factor(year),
    qmod=list(~factor(age)),
    srmod=~1,
    vmod=list(~1, ~1),
    n1mod=~factor(age))
res06 <- residuals(fit06, hke1567, hke1567.idx)
```

The best way to inspect the effect of this model is to zoom into the initial years of the time series.

```{r, echo=FALSE}
r06 <- window(res06, end="2010")
class(r06) <- "a4aFitResiduals"
r05 <- window(res05, end="2010")
class(r05) <- "a4aFitResiduals"
```

Figure \@ref(fig:n1resbyage05) zooms into the previous model, which used an intercept only model for N1, while Figure \@ref(fig:n1resbyage06).

```{r, n1resbyage05, fig.cap="N1 fitted as an intercept only model: 2007 - 2010 residuals by age", echo=FALSE}
plot(r05, auxline="l", by="age")
```

```{r, n1resbyage06, fig.cap=, fig.cap="N1 fitted with an age effect model: 2007 - 2010 residuals by age", echo=FALSE}
plot(r06, auxline="l", by="age")
```

Comparing the two plots it can be seen the effect of modeling abundance in the initial year. Residuals for both catch at age and catchability improved considerably. The following years also improve to different levels.

## The stock recruitment submodel

In this example we'll simply add a model to allow recruitment to vary over time and we'll see how to track potential improvements in the residuals.

```{r}
fit07 <- sca(hke1567, hke1567.idx,
    fmod=~factor(age) + factor(year),
    qmod=list(~factor(age)),
    srmod=~factor(year),
    vmod=list(~1, ~1),
    n1mod=~factor(age))
res07 <- residuals(fit07, hke1567, hke1567.idx)
```

The residuals plot by year are very useful to see the effect of adding a varying stock recruitment model. The year trends present in previous models are not absent. Recruitment variability when not modelled was being picked up by trends in the survey catchability and catch at age. And due to the cohort dynamics underlying the catch at age model, where propagating into other ages' estimates.

```{r, srresbyyear, fig.cap="Stock-recruitment model with year effect residuals by year"}
plot(res07)
```

## The variance submodel

Finally, we're testing the variance submodel, specifically the catch at age variance model. We won't dig into the catchability variance model though. It's common to accept that a scientific survey following a well designed sampling protocol will have equal variance across ages since no preferential areas should be sampled sampled.

The variance model will use a smoother with `k=3`. The expectation is that the variance model will have a U-shape, since younger and older ages are usually less caught and as such estimates of those ages will have larger variances than fully exploited ages.

```{r}
fit08 <- sca(hke1567, hke1567.idx,
    fmod=~factor(age) + s(year, k=10),
    qmod=list(~factor(age)),
    srmod=~s(year, k=10),
    vmod=list(~s(age, k=3), ~1),
    n1mod=~factor(age))
```

We'll use the pearson residuals for this analysis since those are standardized by the predicted variances of the model and not the residual variance itself, like the more common standardized residuals. Figure \@ref(fig:vresbyyear08) shows an improve set of residuals when compared to Figure \@ref(fig:vresbyyear07) which add an intercept only model for the variance model.

```{r}
res07 <- residuals(fit07, hke1567, hke1567.idx, type="pearson")
res08 <- residuals(fit08, hke1567, hke1567.idx, type="pearson")
```

```{r, vresbyyear07, fig.cap="Variance model with intercept only age effect pearson residuals"}
plot(res07)
```

```{r, vresbyyear08, fig.cap="Variance model with age effect pearson residuals"}
plot(res08)
```

To see what's happening with the variance model one can use predict to plot the different models fitted.

```{r, vagepredbyage, fig.cap="Variance models for catch at age", echo=FALSE}
flqs <- FLQuants(mod08=predict(fit08)$vmodel$catch[,"2022"],
    mod07=predict(fit07)$vmodel$catch[,"2022"])
xyplot(data~age, data=flqs, group=qname, type="l", auto.key=T)
```

To see the effect these models have on the estimated quantities one can look at the variance of the estimates:

```{r, vage, fig.cap="Estimates of population abundance with different variance models", echo=FALSE}
flqs <- FLQuants(mod08=catch.n(hke1567+simulate(fit08, nsim=500))[,"2022"], mod07=catch.n(hke1567+simulate(fit07, nsim=500))[,"2022"])
bwplot(data~qname|factor(age),  data=as.data.frame(flqs), scales="free", auto.key=T)
```

## Final comments

- The sequence presented here can be changed and applied in any order the user is interested or prefers to.
- The approach of not allowing year effects in surveys and variance model can be modified if the user prefers to do so.
- The (ab)use of `factor` is for demonstration purposes only. The user is incentivised to explore other model forms, in particular smoothers.







